### 2018-04-06 ident/docs/draft_2.tex
more changes to intro

	In contrast, the implications of regulatory interactions and changes in metabolite concentrations on different characteristics of metabolism can be studied using kinetic models of metabolism \parencite{Saa2017}. These models account for the thermodynamic and regulatory interactions that underly metabolic networks in addition to its stoichiometry \parencite{Link2014}. Kinetic models can not only help us better understand lesser known and understood characteristics of metabolism like bistability \parencite{Kotte2014}, and their role in human health, but can also improve predictions about the impact of engineering design perturbations on metabolism, and propose alternative designs to achieve metabolite production goals \parencite{Khodayari2016}.

@@

	The development of these enzyme kinetic rate laws is based on in vitro observations of enzyme activity. Accordingly, authors have questioned their relevance for gleaning information on the dynamics of metabolism under in vivo conditions, as opposed to in vitro conditions \parencite{Heijnen2013}. This problem is further compounded by the fact that typical parameter values used in kinetic models of metabolism are estimated in vitro, and may not be applicable under in vivo conditions \parencite{Heijnen2005,Smallbone2007}. Moreover, estimation of kinetic parameters based on experimental in vivo data becomes computationally expensive for larger models, and any estimated in vivo parameter estimates are associated with large uncertainties \parencite{Link2014}. Using in vitro, or unreliable in vivo parameter estimates to study the steady state and dynamic responses of metabolic networks, under in vivo conditions, to different perturbations reduces confidence in the model predicted behaviour \parencite{Andreozzi2016a}. Consequently, this hampers the use of these models to gain insight into metabolic network function \parencite{Chakrabarti2013a}, and the increase in prediction uncertainty becomes an obstacle for using the predicted responses as a basis for designing the metabolic networks to achieve any of the aforementioned goals of metabolic engineering and design \parencite{Saa2017}.

	So, the use of heavily parameterized enzyme kinetic rate laws not only differentiates kinetic models from CBMs of metabolism, but is also a major hurdle to the widespread use and adoption of kinetic models \parencite{Berthoumieux2013,Vasilakou2016}. In order to address this problem, authors have proposed two different strategies. In one scenario, authors have looked at the development and use of global optimization algorithms for parameter estimation \parencite{Moles2003}. In another case, in vitro kinetic rate law descriptions are abandoned in favour of approximate and new empirical enzyme kinetic models whose parameters can not only be estimated in a relatively easier fashion, but also lead to kinetic models that generate information that has a higher correlation with experimental data that is observed in vivo \parencite{Heijnen2005,Smallbone2007}.

	However, one of persistent problems is the lingering uncertainty in the parameter estimates and the subsequent predictions that are obtained from these methods. Authors have proposed to alleviate this parameter uncertainty either by using a Monte Carlo approach to kinetic modeling of metabolism \parencite{Srinivasan2015}, or by quantifying this uncertainty in parameter estimates using different methods \parencite{Vanlier2013}. Approaches to parameter estimation from a Bayesian perspective have also been proposed to estimate parameters and quantify the uncertainty in these estimates \parencite{Saa2016}.

	In spite of all these developments, model parameter identifiability, a necessary condition to estimate unique kinetic parameter values from experimental data, is often overlooked \parencite{Ljung1994,Berthoumieux2013}. Briefly, it concerns with the ability to distinguish between any two different parameter values based on differences in the inputs to, and consequently the responses exhibited by the cell. Structural identifiability refers to the dependence of the ability to estimate unique parameter values on model structure, due to inherent dependencies between two or more parameters in the model \parencite{Berthoumieux2013}. Conversely, when the dependence is on the informativeness of the data used for estimation, and not on the structure of the model, it is referred to as practical identifiability \parencite{Ljung1994}.

	Significant work has been done towards the development of methods for structural identification of parameters in kinetic models of metabolism \parencite{Ljung1994,Nikerel2009,Berthoumieux2013,Raue2014}\textcolor{red}{(paper from Rudiyanto Gunawan on model discrimination and sensitivity analysis)}. However, the focus on methods for practical identification of parameters in kinetic models of metabolism and its use for experimental design is fairly recent and limited \parencite{Gadkar2005a,Vanlier2014a,Raue2014}. For instance some of these methods are limited by their applicability to approximate kinetic models only \parencite{Nikerel2009,Berthoumieux2013}, and some perform identifiability within a dynamic context \parencite{Gadkar2005a,Nikerel2009,Vanlier2014a}. Some other methods also suffer from their inability to be applied to kinetic models of large metabolic networks due to the associated computational burden \parencite{Gadkar2005a,Raue2014}\textcolor{red}{(Banga method using FIM for D-optimal design, ??)}.

	In this paper, we propose a scalable methodology that uses available steady state fluxomics, metabolomics and proteomics data to practically identify parameters for each individual reaction in kinetic models of metabolism. We demonstrate how the computer algebra-based method that we have developed can also facilitate the design of experiments that are minimal and informative to generate data required to estimate unique parameter values for all reaction fluxes in a metabolic network. In doing so, we not only propose specific types of and nature of perturbations that will provide the most useful data for parameter estimation, but also test the identifiability of different enzyme kinetic rate laws that are typically used to model fluxes in metbaolic networks.

	 For the purposes of this method we assume that all intracellular metabolite concentrations and fluxes can be measured. We illustrate our methodology to identify parameters and design experiments for estimating parameters in a small metabolic network model of glucoeneogenesis in \textit{Escherichia coli} \parencite{Kotte2014, Srinivasan2017}.  %We also demonstrate the scalability of the proposed methodology to facilitate experimental design by applying it to a relatively larger metabolic network of the human red blood cell hepatocyte.

@@

	\section{Methods}\label{sec:methods}
	\subsection{Definition: Structural and Practical parameter identifiability of kinetic models of metabolism}\label{sec:kinetic_model}
	In kinetic models of metabolism, ordinary differential equations (ode) are used to express the rate of change of metabolite concentrations ($x$) as a function of the reaction fluxes ($v$) in the metabolic network (Equation \ref{eq:kinstoich}). The matrix $\mathbf{S}$ in Equation (\ref{eq:kinstoich}a) defines the stoichiometric relationship between the fluxes and the concentrations of the metabolic network.
	\begin{subequations}\label{eq:kinstoich}
		\begin{align}
		\dot{x} = \mathbf{S}v\\
		v = f(x, \theta, u)
		\end{align}
	\end{subequations}
	The expression for the nonlinear function ($f$) used to describe each reaction flux $v_i$ in $v$, $i={1, 2, ..., n_r}$, in a kinetic model (Equation \ref{eq:kinstoich}b) is dependent on the enzyme kinetic mechanism that is used to model the reaction \parencite{Srinivasan2015}. Accordingly, $f$ is a nonlinear function of the metabolite concentrations ($x$), enzyme kinetic parameters ($\theta$) and other input concentrations ($u$).

@@

	However, not all metabolite concentrations and fluxes in the model (Equation \ref{eq:kinstoich}) change for any random experiment. This makes unambiguous estimation of parameters impossible, either due to the inherent correlation between changes in different concentrations or fluxes, or due to the homeostasis of the concentrations and fluxes under the chosen experimental conditions \parencite{Heijnen2013}. In such scenarios, the need to design experiments to effect a change in and discriminate between changes in different concentrations/fluxes becomes necessary.

	As briefly mentioned in the Introduction, in systems identification, the ability to estimate unique parameter values from available experimental data is governed by the identifiability of these parameters in the model \parencite{Ljung1994,Vanlier2012,Berthoumieux2013,Raue2014}. The identifiability of parameters in linear and nonlinear models can be classified into two categories: structural (or a priori) and practical (or posterior) identifiability.

>>>>

	In contrast, the implications of regulatory interactions and changes in metabolite concentrations on different characteristics of metabolism can be studied using kinetic models of metabolism \parencite{Saa2017}. These models account for changes in metabolite concentrations subject to thermodynamic and regulatory constraints that underly metabolic networks in addition to its stoichiometry \parencite{Link2014}. Kinetic models can not only help us better understand lesser known and understood characteristics of metabolism like bistability \parencite{Kotte2014}, and their role in human health, but can also improve predictions about the impact of engineering design perturbations on metabolism, and propose alternative designs to achieve metabolite production goals \parencite{Khodayari2016}.

@@

	The development of these enzyme kinetic rate laws is based on in vitro observations of enzyme activity. Accordingly, authors have questioned their relevance for gleaning information on the dynamics of metabolism under in vivo conditions, as opposed to in vitro conditions \parencite{Heijnen2005,Heijnen2013}. This problem is further compounded by the fact that typical parameter values used in kinetic models of metabolism are estimated in vitro, and may not be applicable under in vivo conditions \parencite{Heijnen2005,Smallbone2007}. Also, parameters estimated based on in vivo experimental data are associated with large uncertainties \parencite{Link2014}. Using in vitro, or unreliable in vivo parameter estimates to study the steady state and dynamic responses of metabolic networks, under in vivo conditions, to different perturbations reduces confidence in the model predicted behaviour. Consequently, this hampers the use of these models to gain insight into metabolic network function \parencite{Andreozzi2016a,Vasilakou2016}, and the increase in prediction uncertainty becomes an obstacle for using the predicted responses as a basis for designing the metabolic networks to achieve any of the aforementioned goals of metabolic engineering and design \parencite{Saa2017}.

	Although some authors have sought to quantify this uncertainty using different techniques \parencite{Vanlier2013,Andreozzi2016a}, others have proposed to alleviate as well as constrain the uncertainty in parameter estimates and consequent model predictions by using a Monte Carlo approach to kinetic modeling of metabolism that allows for integration of experimentally observed in vivo data \parencite{Srinivasan2015}. Bayesian approaches to improve parameter estimation and quantify estimation uncertainty have also been proposed \parencite{Saa2016}.

	In spite of the development of these methods to quantify parameter estimation uncertainty, model parameter identifiability, a necessary, and sometimes sufficient condition to estimate unique kinetic parameter values from experimental data, is often overlooked \parencite{Ljung1994,Berthoumieux2013}. Briefly, it concerns with the ability to estimate unique values for all model parameters from observed experimental data. In a model, any parameter is said to be structurally or a priori identifiable if its values can be estimated independent of all other model parameters from available experimental data. However, if parameters cannot be estimated independent of each other due to redundant model parameterization, or due to the nonlinear relationship between the model parameters, then the parameters are said to be structurally non-identifiable. Conversely, if the ability to estimate unique parameter values is compromised due to the inability of the available data to capture the requisite information needed to estimate the parameters in the modeled system, and the uncertainty in parameter estimates is unquantifiable, the parameter is said to be practically non-identifiable \parencite{Ljung1994}.

	Authors have proposed to overcome concerns with parameter identifiability by proposing approximate kinetic models of metabolism that utilize empirical enzyme kinetic rate laws whose parameters have physical significance, and are identifiable \parencite{Heijnen2005,Smallbone2007}. Significant work has also been done towards the development of methods for structural identification of parameters in kinetic models of metabolism \parencite{Ljung1994,Nikerel2009,Berthoumieux2013,Raue2014}\textcolor{red}{(paper from Rudiyanto Gunawan on model discrimination and sensitivity analysis)}.

	Methods to improve practical identifiability through a priori experimental design have also been developed, with focus on kinetic models of metabolism \parencite{Gadkar2005a,Vanlier2014a,Raue2014}. Some of these methods are limited by their applicability to approximate kinetic models only \parencite{Nikerel2009,Berthoumieux2013}, while some of them suffer from computational limitations when applied to kinetic models of large metabolic networks \parencite{Gadkar2005a,Raue2014}\textcolor{red}{(Banga method using FIM for D-optimal design, ??)}.

	In this paper, we propose a scalable methodology that uses available steady state fluxomics, metabolomics and proteomics data to test the practical identifiability of parameters for each individual reaction in kinetic models of metabolism. We demonstrate how the computer algebra-based method that we have developed can also facilitate the design of experiments that are minimal and informative to generate data required to estimate unique parameter values for all reaction fluxes in a metabolic network. In doing so, we not only propose the number and types of perturbations that will provide the most useful data for parameter estimation, but also test the identifiability of different enzyme kinetic rate laws that are typically used to model fluxes in metabolic networks.

	For the purposes of this method we assume that all intracellular metabolite concentrations and fluxes can be measured. We illustrate our methodology to identify parameters and design experiments to identify parameters in a small metabolic network model of glucoeneogenesis in \textit{Escherichia coli} \parencite{Kotte2014, Srinivasan2017}.

	%some perform identifiability within a dynamic context \parencite{Gadkar2005a,Nikerel2009,Vanlier2014a} and

	  %We also demonstrate the scalability of the proposed methodology to facilitate experimental design by applying it to a relatively larger metabolic network of the human red blood cell hepatocyte.

@@

	\section{Methods}\label{sec:methods}
	\subsection{Parameter estimation for kinetic models of metabolism}\label{sec:kinetic_model}
	In kinetic models of metabolism, ordinary differential equations (ODE) are used to express the rate of change of metabolite concentrations ($x$) as a function of the reaction fluxes ($v$) in the metabolic network (Equation \ref{eq:kinstoich}). The matrix $\mathbf{S}$ in Equation (\ref{eq:kinstoich}a) defines the stoichiometric relationship between the fluxes and the concentrations of the metabolic network.
	\begin{subequations}\label{eq:kinstoich}
		\begin{align}
		\dot{x} = \mathbf{S}v\\
		v = f(x, \theta, u)
		\end{align}
	\end{subequations}
	The expression for the nonlinear function ($f$) used to describe each reaction flux $v_i$ in $v$, $i={1, 2, ..., n_r}$, in a kinetic model (Equation \ref{eq:kinstoich}b) is dependent on the enzyme kinetic mechanism that is used to model the reaction \parencite{Srinivasan2015}. Accordingly, $f$ is a nonlinear function of the metabolite concentrations ($x$), enzyme kinetic parameters ($\theta$) and other input concentrations ($u$).

@@

	However, not all metabolite concentrations and fluxes in the model (Equation \ref{eq:kinstoich}) change for any random experiment. This makes unambiguous estimation of parameters impossible, either due to the inherent correlation between changes in different concentrations or fluxes, or due to the homeostasis of the concentrations and fluxes under the chosen experimental conditions \parencite{Heijnen2013}. In such scenarios, the need to design experiments to effect a change in, and discriminate between changes in different concentrations/fluxes becomes necessary.

	\subsection{Structural and practical identifiability of parameters in kinetic models}	
	As briefly mentioned in the Introduction, the ability to estimate unique parameter values from available experimental data is governed by the identifiability of these parameters in the model \parencite{Ljung1994,Vanlier2012,Berthoumieux2013,Raue2014}. The identifiability of parameters in linear and nonlinear models can be classified into two categories: structural (or a priori) and practical (or posterior) identifiability.
